Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-11 paper/type-inference.tex
Edits to type inference.

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. Likewise, given the region-annotated definition of
\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
is \footnote{In our exposition, we assume that classes $\C{A}$ and
$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
$\ObjZ$, they accept one region parameter - the allocation region of
their objects.}\footnote{We abuse arrow notation to also represent
types of methods, but unlike function types, there is no allocation
region annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
variables, and $\varphi_0$ is a fresh predicate variable denoting
unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
fresh. For a class, a template is computed in two steps. In the first
step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following (method definitions
are left intact while templatizing, hence they are elided below):
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5}$() { $...$ }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$.

@@

Our constraint generation algorithm generates three kinds of
constraints:
\begin{itemize}
\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to the set ($\rhoenv$) of regions in scope,
\item Well-formedness constraints of form
$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas
over region variables ($\rhoenv$) in scope, and
\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
extended with predicate variables and \emph{pending
substitutions}\footnote{we borrowed this terminology
from~\cite{ltpldi08}}:
\begin{smathpar}
\begin{array}{lcl}
\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
    \ALT F(\varphi) \ALT \phi \conj \phi\\
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
\end{itemize}
A pending substitution ($F$) is a substitution function over region
variables/identifiers. They represent the substitutions that need to
be carried out when a predicate variable ($\varphi$) is replaced by a
concrete formula in a validity constraint. For instance, in the
validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.

% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
In general, validity constraints generated by our algorithm assume one
of the following two forms:
\begin{center}
\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {\phicstr} \qquad
   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {F_j(\varphi_j)}
\)
\end{center}
Where $\varphi_i$'s denote the unknown preconditions of the class and
the method under elaboration. If the constraint is generated while
checking the well-formedness of a type or elaborating an expression,
then $\varphi_j$'s denote the unknown preconditions of classes and
methods that were used in that type or expression. Each use of a
(region-polymorphic) class or a method has to instantiate region
parameters, resulting in pending substitutions ($F_j$). Formulas
$\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
variables and pending substitutions. While $\phictxt$ captures
relationships that are \emph{known} to hold between concrete region
identifiers (i.e., $\rgn$'s) when the constraint was generated,
$\phicstr$ captures relationships that are \emph{required} to hold
among region varibles (i.e., $\rho$'s), or relationships between
region variables and identifiers. Each region variable occuring in
$\phicstr$ has an associated well-formedness constraint, which
specifies its unification domain. The unification domain of a
constraint is the union of unification domains of all region variables
occuring in the constraint.

\paragraph{Constraints Example 1} The original source of all
outlives constraints is the validity constraint $C_3$ in $\elabClass$
function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
class template from \S\ref{sec:fb-templatization}, the constraint is
as following:
\begin{center}
  \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
  \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
\end{center}

\paragraph{Constraints Example 2} Consider a contrived method
$\C{alt}$ that accepts an argument $\C{p}$ of type
$\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
of type $A$. It then assigns one object to $\C{p.fst}$ and other to
$\C{p.snd}$, and calls itself recursively without ever terminating.
Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
alternated between recursive calls. The region-annotated definition of
$\C{alt}$ is shown below:
\begin{codejava}
unit alt<$\rhoalloc$,$\rho_{0-4}$ | $\varphi_0$>(Pair<A,A><$\rho_0$,$\rho_1$,$\rho_2$> p,
                          A<$\rho_3$> x, A<$\rho_4$> y) {
  return p.fst = x; p.snd = y; 
         this.alt<$\rhoalloc$,$\rho_1$,$\rho_2$,$\rho_4$,$\rho_3$>(A,y,x)
}
\end{codejava}
Note that, to avoid clutter, we have already resolved appropriate
region arguments to the recursive call, instead of introducing new
region variables and generating equality constraints on them. Rest of
the validity constraints generated while elaborating $\C{alt}$ to the
above region-annotated definition are shown below:
\begin{center}
\(\isvalid{\varphi_0}{\rho_1 \outlives \rho_0 \conj \rho_2 \outlives \rho_0}\)
$\quad$
\(\isvalid{\varphi_0}{\rho_1 = \rho_3}\)
$\quad$
\(\isvalid{\varphi_0}{\rho_2 = \rho_4}\)\\
\(\isvalid{\varphi_0}{[\rho_3/\rho_4][\rho_4/\rho_3]\varphi_0}\)
\end{center}
The first constraint is generated by $\typeOk$ on the type of $\C{p}$.
Second and third are generated by the the assignment expressions, and
the last constraint is generated from the recursive call.

>>>>

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. Likewise, given the region-annotated definition of
\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
is \footnote{In our exposition, we assume that classes $\C{A}$ and
$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
$\ObjZ$, they accept one region parameter - the allocation region of
their objects.}\footnote{We abuse arrow notation to also represent
types of methods, but unlike function types, there is no allocation
region annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
variables, and $\varphi_0$ is a fresh predicate variable denoting
unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
fresh. For a class, a template is computed in two steps. In the first
step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following :
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
    super(); this.fst=fst; this.snd=snd; 
  }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$.

@@

Our constraint generation algorithm generates three kinds of
constraints:
\begin{itemize}
\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to the set ($\rhoenv$) of regions in scope,
\item Well-formedness constraints of form
$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas
over region variables ($\rhoenv$) in scope, and
\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
extended with predicate variables and \emph{pending
substitutions}\footnote{we borrowed this terminology
from~\cite{ltpldi08}}:
\begin{smathpar}
\begin{array}{lcl}
\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
    \ALT F(\varphi) \ALT \phi \conj \phi\\
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
\end{itemize}
A pending substitution ($F$) is a substitution function over region
variables/identifiers. They represent the substitutions that need to
be carried out when a predicate variable ($\varphi$) is replaced by a
concrete formula in a validity constraint. For instance, in the
validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.
% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
% In general, validity constraints generated by our algorithm assume one
% of the following two forms:
% \begin{center}
% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {\phicstr} \qquad
%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {F_j(\varphi_j)}
% \)
% \end{center}
% Where $\varphi_i$'s denote the unknown preconditions of the class and
% the method under elaboration. If the constraint is generated while
% checking the well-formedness of a type or elaborating an expression,
% then $\varphi_j$'s denote the unknown preconditions of classes and
% methods that were used in that type or expression. Each use of a
% (region-polymorphic) class or a method has to instantiate region
% parameters, resulting in pending substitutions ($F_j$). Formulas
% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
% variables and pending substitutions. While $\phictxt$ captures
% relationships that are \emph{known} to hold between concrete region
% identifiers (i.e., $\rgn$'s) when the constraint was generated,
% $\phicstr$ captures relationships that are \emph{required} to hold
% among region varibles (i.e., $\rho$'s), or relationships between
% region variables and identifiers. Each region variable occuring in
% $\phicstr$ has an associated well-formedness constraint, which
% specifies its unification domain. The unification domain of a
% constraint is the union of unification domains of all region variables
% occuring in the constraint.

\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
template from \S\ref{sec:fb-templatization}. Following constraints are
generated during its elaboration (Constrains are identified with
$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
\rhoenv_0$, where $\rhoenv_0 = \{\rhoalloc_0,\rho_{0-4}\}$, have been
elided): 
\begin{smathpar}
\begin{array}{l}
  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \spc
  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
\end{array}
\end{smathpar}

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts an argument $\C{p}$ of type
$\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
of type $A$. It then assigns one object to $\C{p.fst}$ and other to
$\C{p.snd}$, and calls itself recursively without ever terminating.
Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
alternated between recursive calls. Elaboration phase elaborates
$\C{alt}$ to the following region-annotated definition\footnote{The
original definition of \C{alt} can be obtained by erasing all the
region annotations from the elaborated version}\footnote{The
elaboration actually uses new region variables as parameters to the
recursive call, and then generates constraints that unify them with
the ones shown. We coalesced both steps to simplify presentation.}:
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
                          A<$\rho_9$> x, A<$\rho_{10}$> y) {
  p.fst = x; p.snd = y; 
  this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-10}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{6} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{7} \isvalid{\varphi_2}{\rho_7 \outlives \rho_6 \conj \rho_8
  \outlives \rho_6} \\
\csid{8} \isvalid{\varphi_2}{\rho_9 = \rho_7} \qquad
\csid{9} \isvalid{\varphi_2}{\rho_{10} = \rho_8} \\
\csid{10} \isvalid{\varphi_2}{[\rho_9/\rho_{10}][\rho_{10}/\rho_9]\varphi_2}
\end{array}
\end{smathpar}
% The first constraint is generated by $\typeOk$ on the type of
% $\C{p}$.  Second and third are generated by the the assignment
% expressions, and the last constraint is generated from the recursive
% call.
